---
title: "R for DS Chapter 5 - Answers"
output: html_notebook
---
```{r}
library(tidyverse)
library(nycflights13)
```

##5.2.4 EXERCISES
####1. Find all flights that
```{r}
ex1 <- flights$arr_delay >= 120
ex2 <- flights$dest %in% c("IAH","HOU")
ex3 <- flights$carrier %in% c("AA","UA","DL")
ex4 <- flights$month %in% c(7,8,9)
ex5 <- (flights$arr_delay > 120) & (flights$dep_delay <= 0)
ex6 <- (flights$arr_delay >= 60) & (flights$dep_delay - flights$arr_delay > 30)
ex7 <- flights$hour <= 6 | flights$hour == 24
```
####2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?
    No.
####3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?
    These are cancelled flights. Arrival and flight related columns are also missing.
```{r}
filter(flights, is.na(flights$dep_time))
```
####4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)
    These are operations which have the same results for every possible input.

##5.3.1 EXERCISES
####1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).
    First order by is.na() desc for putting the column NA to the top than again arrange by the same column by preferance.
```{r}
  ex1 <- tail(arrange(flights,desc(is.na(flights$dep_delay)), dep_delay))
  ex1
```
####2. Sort flights to find the most delayed flights. Find the flights that left earliest.
ex1 gives the top6 delayed flights and ex2 gives the earliest departed flights.
```{r}
  ex2 <- tail(arrange(flights,desc(is.na(flights$dep_delay)), desc(dep_delay)))
  ex2
```
####3. Sort flights to find the fastest flights.
    Sort by air_time and take head with n.
  
```{r}
ex3 <- arrange(flights, air_time)
ex3
```
####4. Which flights travelled the longest? Which travelled the shortest?
    Shortest travelled flights are given in ex3. In minutes.
    
```{r}
ex4 <- arrange(flights, desc(air_time))
ex4
```
##5.4.1 EXERCISES
####1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

```{r}
# The most obvious
a <- select(flights, dep_time, dep_delay,arr_time, arr_delay)
# starts with
select(flights, starts_with("dep"), starts_with("arr"))
# contains
select(flights, contains("delay"), arr_time, dep_time)
```
####2. What happens if you include the name of a variable multiple times in a select() call?
It takes it only once.

```{r}
select(flights, dep_time, dep_time)
```
####3. What does the one_of() function do? Why might it be helpful in conjunction with this vector?
    one_of() function allows you to select from a charachter vector.
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
```
####4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
The select functions are not case sensitive by default by ignore.case argument can be changed to FALSE.
```{r}
select(flights, contains("TIME"))
```
Outputs an empty tibble.
```{r}
empty <- select(flights, contains("TIME",ignore.case = FALSE ))
empty
```
####5.5.2 EXERCISES
##1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r}
f2 <- transmute(flights,
    dep_time,
    dep_time_minutes = (dep_time %/% 100)*60 + (dep_time %% 100) 
)

f3 <- transmute(flights,
    sched_dep_time,
    sched_dep_time_minutes = (sched_dep_time %/% 100)*60 + (sched_dep_time %% 100) 
)
```
####2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?

```{r}
flights_sml <- select(flights, dep_time, arr_time, air_time)
mutate(flights_sml, 
       arr_min = (arr_time %/% 100)*60 + (arr_time %% 100),
       dep_min = (dep_time %/% 100)*60 + (dep_time %% 100),
       trip_min = arr_min - dep_min
       )
```
####3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?
    dep_delay = dep_time - sched_dep_time
```{r}
select(flights, dep_time, sched_dep_time, dep_delay)
```

####4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? Carefully read the documentation for min_rank().
```{r}
#TIES WILL BE BROKEN WHILE FIRST WILL GET SMALLER RANK
head(arrange(flights,desc(row_number(flights$dep_delay))), 10)

# TIES WILL ALL GET THE SAME SMALLEST RANK
head(arrange(flights,desc(min_rank(flights$dep_delay))), 10)
```
####5. What does 1:3 + 1:10 return? Why?
Gives error, beacuse length of shorter object must be multiple of the longer one.
```{r}
1:3 + 1:10
```
####6. What trigonometric functions does R provide?
    Degree is in radians.
    
    cos(x)
    sin(x)
    tan(x)
    
    acos(x)
    asin(x)
    atan(x)
    atan2(y, x)
    
    cospi(x)
    sinpi(x)
    tanpi(x)
    

```{r}
x <- seq(-3, 7, by = 1/8)
tx <- cbind(x, cos(pi*x), cospi(x), sin(pi*x), sinpi(x),
               tan(pi*x), tanpi(x), deparse.level=2)
op <- options(digits = 4, width = 90) # for nice formatting
head(tx)

```
##5.6.7 EXERCISES
####1. Brainstorm at least 5 different ways to assess the typical delay characteristics of a group of flights. Consider the following scenarios:
      - A flight is 15 minutes early 50% of the time, and 15 minutes late 50% of the time.
      - A flight is always 10 minutes late.
      - A flight is 30 minutes early 50% of the time, and 30 minutes late 50% of the time.
      - 99% of the time a flight is on time. 1% of the time it’s 2 hours late.
        Which is more important: arrival delay or departure delay?
        
        
```{r}
###BE SURE EACH FLIGHT HAD AIR TIME
not_ancelled <- flights %>%
  filter(!is.na(air_time))

###ARRIVAL IS MORE IMPORTANT SO SORT FIRST ON THAT
not_cancelled %>%
  group_by(tailnum) %>%
  mutate(
    count = n(),
    median_arr_delay = median(arr_delay),
    median_dep_delay = median(dep_delay)
  ) %>%
  filter(count > 30) %>%
  arrange(median_arr_delay, median_dep_delay)
  
```
      From the table we see that N479AA no. plane arrives early more than 20 mins for the 50% of the time, departs early more than 3 mins for the 50% of the time.
      
      
```{r}
###ARRIVAL IS MORE IMPORTANT SO SORT FIRST ON THAT
not_cancelled %>%
  group_by(tailnum) %>%
  summarise(
    count = n(),
    p_15_early_arr = mean(arr_delay < -15),
    p_15_dep_arr = mean(dep_delay < -15)
  ) %>%
  filter(p_15_early_arr > 0.5 | p_15_dep_arr > 0.5) %>%
  filter(count > 30) %>%
  arrange(desc(p_15_early_arr), desc(p_15_dep_arr))
  
```

      Here we have the flights that made at least 30 flights within a year and have arrive or depart earlier than 15 mins for more than 50% of the time. Arrival is more important, since departing can be compensated in air. 



```{r}
###MORE THAN 10 MINS LATE
not_cancelled %>%
  group_by(tailnum, origin, dest) %>%
  summarise(
    count = n(),
    arr_delay_10_c = sum(arr_delay > 10),  
    arr_delay_10_p = mean(arr_delay > 10),  
    dep_delay_10_c = sum(dep_delay > 10),
    dep_delay_10_p = mean(dep_delay > 10)  
    ) %>%
  filter(count > 20) %>%
  arrange(desc(arr_delay_10_p))
  
```

      Flights from LaGuardia to Atlanta arrive with a delay more than 10 minutes most of the time.

```{r}
###EXACT 10 MINS OF LATE PROBABILITY
not_cancelled %>%
  group_by(tailnum) %>%
  summarise(
    count = n(),
    exact_10 = mean(arr_delay == 10)
  ) %>% 
  filter(count > 10) %>%
  arrange(desc(exact_10))


```

```{r}
### SELECT SEATTLE FOR 30 MINS APPROACH
### BEST FLIGHTS UNDER 30 MINS EARLY OR LATE CRITERIA SORTED
### ARRIVAL IS MORE IMPORTANT DURING SORT
not_cancelled %>%
  group_by(tailnum) %>%
  mutate(
    count = n(),
    arr_30_early = mean(arr_delay < -30),
    dep_30_early = mean(dep_delay < -30),
    arr_30_late  = mean(arr_delay > 30),
    dep_30_late = mean(dep_delay > 30)
  ) %>%
  filter(count > 20) %>%
  arrange(desc(arr_30_early), desc(dep_30_early), arr_30_late, dep_30_late) %>%
  select(dest)

```


