# Chapter 10 Tibbles
```{r}
library(tidyverse)
```

## 10.5 EXERCISES

#### 1. How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).
  We can check the class of the data by using class().
```{r}
class(mtcars)
```

#### 2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?
  As it can be inspect by the following codes, `tbl$x` causes an error since tibbles don't return TRUE by partial matching. We must run `tbl$xyz` to get the exact column. This might cause us a problem when there are to columns starting with x and we intend to get a specific one, not catching an error won't be good in this case.

```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

```{r}
tbl <- as.tibble(df)
tbl$x
tbl[, "xyz"]
tbl[, c("abc", "xyz")]
```
```{r}
tbl$xyz
```

#### 3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?

```{r}
var <- "mpg"
mtcars_tbl <- as.tibble(mtcars)
mtcars_tbl[var] # returns a tibble
mtcars_tbl[[var]] # returns a vector
# same above using pipe
mtcars_tbl %>% 
  .[var]
mtcars_tbl %>%
  .[[var]]
```
#### 4. Practice referring to non-syntactic names in the following data frame by
* Extracting the variable called 1.
* Plotting a scatterplot of 1 vs 2.
* Creating a new column called 3 which is 2 divided by 1.
* Renaming the columns to one, two and three.
  R has some syntatic rules when assigning names. But by using "``" we can assign non syntactic names.
```{r}
# column 2 takes column 1 multiplies each element by 2 then creates a normal (gaussian) random variables vector of size of the 1 column, finally add these to vectors to form a new column called 2.
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
annoying
```
```{r}
class(annoying)
annoying$`1`
```
```{r}
annoying %>%
ggplot(aes(x = `1`, y = `2`))+
  geom_point()
```
```{r}
annoying <- annoying %>%
  mutate(
    `3` = `2`/`1`
  )
```

```{r}
names(annoying) = c("one", "two", "three")
annoying
```
  To those who are familiar with pandas. Enframe and deframe are similar to pd.pivot/pd.unstack and pd.stack. What they do are simply pivoting and stacking data. If you have a long list with names, it can transform it into a dataframe, or vice-versa.
```{r}
stacked_tbl <- tibble(
  groups = as.vector(c(rep("A", 10), rep("B", 10), rep("C", 10))),
  values = rnorm(30)
)
```

```{r}
list_form <- deframe(stacked_tbl)
list_form
```

```{r}
df_form <- enframe(list_form)
df_form
```
#### 6. What option controls how many additional column names are printed at the footer of a tibble?
  One can see the set options by running options(). 

Tibble Package options
  * Display options for tbl_df, used by trunc_mat() and (indirectly) by print.tbl().
  * tibble.print_max Row number threshold: Maximum number of rows printed. Set to Inf to
always print all rows. Default: 20.
  * tibble.print_min Number of rows printed if row number threshold is exceeded. Default: 10.
  * tibble.width Output width. Default: NULL (use width option).
  * tibble.max_extra_cols Number of extra columns printed in reduced form. Default: 100.

```{r}
# No more than 5 column at the end
options(tibble.max_extra_cols = 5)
```



