# Chapter 12 - Tidy Data
```{r}
library(tidyverse)
```

## 12.2.1 EXERCISES

#### 1. Using prose, describe how the variables and observations are organised in each of the sample tables.

```{r}
table1
```
* This table is in tidy data form. Each variable has its own column.

```{r}
table2
```
* This table is not tidy. Type column consist of two variables which should be broken into two separate columns.

```{r}
table3
```
* This table is not tidy as well since we know that rate consists of two different information. But if rate column was given with decimal points without any prior knowledge this data would be tidy.

```{r}
table4a
```

```{r}
table4b
```
* These two tables are again problematic in the means of tidiness since columns should be desciptive names of the observation vector and each column should be unique. These table violate every rule out there. Column names which are years should be another column, cases in table4a should be another column and counts in table4b should be another one. After this these two should be merged by country + year id.

#### 2. Compute the rate for table2, and table4a + table4b. You will need to perform four operations:

* Extract the number of TB cases per country per year.
* Extract the matching population per country per year.
* Divide cases by population, and multiply by 10000.
* Store back in the appropriate place.
* Which representation is easiest to work with? Which is hardest? Why?



```{r}
cases <- table2 %>%
  filter(type == "cases")
populations <- table2 %>%
  filter(type == "population")
merged <- cbind(cases, populations)
merged$rates= (cases$count / populations$count)*1000
merged

```


```{r}
rates_99 <- (table4a$`1999` / table4b$`1999`)*1000
rates_00 <- (table4a$`2000`/ table4b$`2000`)*1000
new <- table4a['country']
new$rates_00 = rates_00
new$rates_99 = rates_99
new
```
#### 3. Recreate the plot showing change in cases over time using table2 instead of table1. What do you need to do first?
  First we need to filter data by cases. Then plot it by count as y variable.
```{r}
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

# below gives the same plot
table2 %>%
  filter(type == "cases") %>%
  ggplot(aes(year, count)) + 
  geom_line(aes(group = country), colour="grey50")+ 
  geom_point(aes(colour = country))
```

## 12.3.3 EXERCISES

#### 1. Why are gather() and spread() not perfectly symmetrical? Carefully consider the following example:
  These two functions are not perfectly symmetrical because gather requires each column name that will stacked together. 

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2016, 2016),
  half  = c(   1,    2,     1,    2),
  return = c(1.88, 0.59, 0.92, 0.17)
)
stocks
stocks %>% 
  spread(year, return) %>% 
  gather("year", "return", `2015`:`2016`)
```
  convert() argument in spread check each newly created column and their string values in order to convert them to appropriate formats. It is useful whne value column before spread function contains mixed data types.

```{r}
# spread with convert TRUE
tbl <- tibble(
  number = c(1,1,2),
  type = c("date", "integer", "date"),
  value = c("12/08/1993", "24", "16/10/1995")
)
spd <- spread(tbl, key = type, value = value, convert = TRUE) #check column types of spd
spd
```
  convert() argument in gather check type of each key/column name before gathering them into one column. If TRUE will automatically run type.convert on the key column. This is useful if the column names are actually numeric, integer, or logical.


```{r}
tbl <- tibble(
  '2017' = c(10,20),
  '2018' = c(30,40), 
  '2019' = c(50,60),
  'x' = c(1,2)
)
tbl
gtr <- gather(tbl, `2017`, `2018`, `2019`, key = "year", value = "y", convert = TRUE)
gtr # type of year column is coerced to be integer
```

#### 2. Why does this code fail?
  1999 and 2000 are not positions of the columns hence we should tell the function they are column names by using backticks.

```{r}
table4a %>% 
  gather(1999, 2000, key = "year", value = "cases")
```
```{r}
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```
#### 3. Why does spreading this tibble fail? How could you add a new column to fix the problem?
  A key can't have multiple columns where age has 2 different values for Phillip Woods. We can add a middle name column and tell that they are different people.

```{r}
people <- tribble(
  ~name,             ~key,    ~value,
  #-----------------|--------|------
  "Phillip Woods",   "age",       45,
  "Phillip Woods",   "height",   186,
  "Phillip Woods",   "age",       50,
  "Jessica Cordero", "age",       37,
  "Jessica Cordero", "height",   156
)

spread(people, key, value)

```
```{r}
people <- tribble(
  ~name,             ~key,    ~value, ~middle,
  #-----------------|--------|------|----------|
  "Phillip Woods",   "age",       45, "J",
  "Phillip Woods",   "height",   186, "J",
  "Phillip Woods",   "age",       50, "M",
  "Jessica Cordero", "age",       37, "",
  "Jessica Cordero", "height",   156, ""
)

spread(people, key, value)
```

#### 4. Tidy the simple tibble below. Do you need to spread or gather it? What are the variables?

```{r}
preg <- tribble(
  ~pregnant, ~male, ~female,
  "yes",     NA,    10,
  "no",      20,    12
)

temp <- gather(preg, key = "gender", value = "count", male, female)
tidy <- spread(temp, key = pregnant, value = count)
```
## 12.4.3 EXERCISES

#### 1. What do the extra and fill arguments do in separate()? Experiment with the various options for the following two toy datasets.
  fill:
    If sep is a character vector, this controls what happens when there are not enough pieces. There are three valid options:
  * "warn" (the default): emit a warning and fill from the right
  * "right": fill with missing values on the right
  * "left": fill with missing values on the left
  extra:
    If sep is a character vector, this controls what happens when there are too many pieces. There are three valid options:
  * "warn" (the default): emit a warning and drop extra values.
  * "drop": drop any extra values without a warning.
  * "merge": only splits at most length(into) times
```{r}
tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
  separate(x, c("one", "two", "three"), extra = "merge")

tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
  separate(x, c("one", "two", "three"), fill = "right")
```

#### 2. Both unite() and separate() have a remove argument. What does it do? Why would you set it to FALSE?
    remove: If TRUE, remove input column from output data frame. You would set it to FALSE if you don't want to lose the original column(s).
    
#### 3. Compare and contrast separate() and extract(). Why are there three variations of separation (by position, by separator, and with groups), but only one unite?
  separate: Given either regular expression or a vector of character positions, separate() turns a single character column into multiple columns.
  extract: Given a regular expression with capturing groups, extract() turns each group into a new column. If the groups don't match, or the input is NA, the output will be NA.

## 12.5.1 EXERCISES

#### 1. Compare and contrast the fill arguments to spread() and complete().
  For spread(): If set, missing values will be replaced with this value. Note that there are two types of missingness in the input: explicit missing values (i.e. NA), and implicit missings, rows that simply aren't present. Both types of missing value will be replaced by fill.
  For complete(): A named list that for each variable supplies a single value to use instead of NA for missing combinations  

```{r}
df <- data_frame(
  group = c(1:2, 1),
  item_id = c(1:2, 2),
  item_name = c("a", "b", "b"),
  value1 = 1:3,
  value2 = 4:6
)
df
complete(df, group, nesting(item_id ,item_name), fill = list(value1 = 0, value2 = 0))
```

#### 2. What does the direction argument to fill() do?
  Direction in which to fill missing values. Currently either "down" (the default) or "up". Like ffill and bfill in pandas.

## 12.6 CASE STUDY

```{r}
who1 <- who%>% 
  gather(new_sp_m014:newrel_f65, key = "key", value = "cases", na.rm = TRUE)

who2 <- who1 %>%
  mutate(
    key = stringr::str_replace(key, "newrel", "new_rel")
  ) 

who3 <- who2 %>%
  separate(key, c("new", "type", "sexage"), sep = "_") %>%
  separate(sexage, c("sex", "age_group"), sep = 1) %>%
  select(-new, -iso2, -iso3)
```


## 12.6.1 EXERCISES

#### 1. In this case study I set na.rm = TRUE just to make it easier to check that we had the correct values. Is this reasonable? Think about how missing values are represented in this dataset. Are there implicit missing values? What’s the difference between an NA and zero?

  I don't think it is very reasonable to drop NAs since they also include information in themselves. Missing values are present when a group experiment/test is absent. 
  
  Running the below shows us that there are many implicit missing values.
```{r}
who3 %>%
  complete(country, year, type, age_group, sex) %>%
  View()
```

#### 2. What happens if you neglect the mutate() step? (mutate(key = stringr::str_replace(key, "newrel", "new_rel")))
  newrel stays as it is and during separate phase new gets newrel, type gets sexage value and sexage gets NA.
```{r}
who %>%
  gather(code, value, new_sp_m014:newrel_f65, na.rm = TRUE) %>% 
  separate(code, c("new", "var", "sexage")) %>%
  View()
```
#### 3. I claimed that iso2 and iso3 were redundant with country. Confirm this claim.
  This gives the unique mapping values for each country.
```{r}
who %>%
  select(country, iso2, iso3) %>%
  group_by(country) %>%
  summarise(
    iso2_value = unique(iso2),
    iso3_value = unique(iso3)
  )
```

#### 4. For each country, year, and sex compute the total number of cases of TB. Make an informative visualisation of the data.

```{r}
summary <- who3 %>%
  group_by(country, year, sex) %>%
  summarise(
    total_cases = sum(cases, na.rm = TRUE)
  )

macro <- summary %>%
  group_by(year, sex) %>%
  summarise(
    total = sum(total_cases)
  )

macro %>%
  ggplot(aes(x = year, y = total, color = sex))+
  geom_smooth(se = FALSE) + 
  geom_point()+
  labs(title = "Sum of total TB cases by year")

```

```{r}
summary %>%
  filter(sex == "m" & year > 1995) %>%
  ggplot(aes(x = year, y  = total_cases, color = country))+
  geom_point(show.legend = FALSE)+
  geom_line(show.legend = FALSE)+
  labs(title= "For Male Population")

summary %>%
  filter(sex == "f" & year > 1995) %>%
  ggplot(aes(x = year, y  = total_cases, color = country))+
  geom_point(show.legend = FALSE)+
  geom_line(show.legend = FALSE)+
  labs(title= "For Female Population")
```


