# Chapter 12 - Data Import
```{r}
library(tidyverse)
```

## 11.2.2 EXERCISES

#### 1. What function would you use to read a file where fields were separated with “|”?

    read_delim(file = ..., delim = "|")
    
#### 2. Apart from file, skip, and comment, what other arguments do read_csv() and read_tsv() have in common?

    col_names, col_types, locale, na, quoted_na, trim_ws, n_max, guess_max, progress
    
#### 3. What are the most important arguments to read_fwf()?
    
    col_positions is the most important argument since the function needs to know length of every field.
    
#### 4. Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like " or '. By convention, read_csv() assumes that the quoting character will be ", and if you want to change it you’ll need to use read_delim() instead. What arguments do you need to specify to read the following text into a data frame?

```{r}
read_delim("x,y\n1,'a,b'", delim = ",",quote = "'", col_names = FALSE)
```
#### 5. Identify what is wrong with each of the following inline CSV files. What happens when you run the code?

```{r}
read_csv("a,b\n1,2,3\n4,5,6") # drops 3rd column, sets the number of columns by looking at the first line
read_csv("a,b\n1,2,3\n4,5,6", col_names = c("x1", "x2", "x3")) 

read_csv("a,b,c\n1,2\n1,2,3,4") # drops 3rd column, sets the number of columns by looking at the first line
read_csv("a,b,c\n1,2\n1,2,3,4", col_names = c("x1", "x2", "x3", "x4"))

read_csv("a,b\n\"1", col_names = FALSE)

read_csv("a,b\n1,2\na,b") # col_names should be set FALSE
read_csv("a,b\n1,2\na,b", col_names = FALSE)

read_csv("a;b\n1;3") # change delim to ;
read_delim("a;b\n1;3", delim = ";", col_names = FALSE)
```

## 11.3.5 Exercises

#### 1. What are the most important arguments to locale()?

    locale(date_names = "en", date_format = "%AD", time_format = "%AT",
      decimal_mark = ".", grouping_mark = ",", tz = "UTC",
      encoding = "UTF-8", asciify = FALSE)
    
    
#### 2. What happens if you try and set decimal_mark and grouping_mark to the same character? What happens to the default value of grouping_mark when you set decimal_mark to “,”? What happens to the default value of decimal_mark when you set the grouping_mark to “.”?
* Error: `decimal_mark` and `grouping_mark` must be different
* Default becomes "."
* Default becomes ","
  
#### 3. I didn’t discuss the date_format and time_format options to locale(). What do they do? Construct an example that shows when they might be useful.

* They set default local date and time formats
* This can be useful if the type of datetime format always same


#### 4. If you live outside the US, create a new locale object that encapsulates the settings for the types of file you read most commonly.

```{r}
locale(date_format = "%d-%m-%Y", time_format = "%H:%M:%S", decimal_mark = ".", grouping_mark = ",",encoding = "utf8")
```
#### 5. What’s the difference between read_csv() and read_csv2()?

  read_csv2() uses ; for separators, instead of ,. This is common in European countries which use , as the decimal separator.

#### 6. What are the most common encodings used in Europe? What are the most common encodings used in Asia? Do some googling to find out.
European 
* ISO 8859-1 Western Europe
* ISO 8859-2 Western and Central Europe
* ISO 8859-3 Western Europe and South European (Turkish, Maltese plus Esperanto)
* ISO 8859-4 Western Europe and Baltic countries (Lithuania, Estonia, Latvia and Lapp)
* ISO 8859-9 Western Europe with amended Turkish character set
* ISO 8859-10 Western Europe with rationalised character set for Nordic languages, including complete Icelandic set

Asian
* KOI8-R, KOI8-U, KOI7
* MIK
* ISCII
* TSCII
* VISCII
* JIS X 0208 is a widely deployed standard for Japanese character encoding that has several encoding forms.
Shift JIS (Microsoft Code page 932 is a dialect of Shift_JIS)
* EUC-JP
* ISO-2022-JP
* JIS X 0213 is an extended version of JIS X 0208.
* Shift_JIS-2004
* EUC-JIS-2004
* ISO-2022-JP-2004
* Chinese Guobiao
* GB 2312
* GBK (Microsoft Code page 936)
* GB 18030
* Taiwan Big5 (a more famous variant is Microsoft Code page 950)
* Hong Kong HKSCS
* Korean
* KS X 1001 is a Korean double-byte character encoding standard
* EUC-KR
* ISO-2022-KR

#### 7. Generate the correct format string to parse each of the following dates and times:

```{r}
d1 <- "January 1, 2010"
d2 <- "2015-Mar-07"
d3 <- "06-Jun-2017"
d4 <- c("August 19 (2015)", "July 1 (2015)")
d5 <- "12/30/14" # Dec 30, 2014
t1 <- "1705"
t2 <- "11:15:10.12 PM"


parse_date(d1, "%B %d, %Y")
parse_date(d2, "%Y-%b-%y")
parse_date(d3, "%d-%b-%Y")
parse_vector(d4,collector = col_date(format = "%B %d (%Y)"))
parse_date(d5, "%m/%d/%y")
parse_time(t1, "%H%M")
parse_time(t2,"%H:%M:%OS %p")
```

